---
title: "Sexual Dimorphism in Penguin Bill Length Across Species"
author: "Josh Abel-Nwachukwu"
date: "2025-09-28"
output: pdf_document
bibliography: references.bib
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center",
  out.width = "75%"   #scale for  images
)
library(dplyr)
library(knitr)

```

# **Introduction**

Sexual dimorphism, or biological differences between males and females of the same species, is a widespread phenomenon with important implications for ecology, evolution, and biodiversity [@tsuji2020]. In birds, variation in bill morphology often reflects ecological specialization, sexual selection, or niche differentiation between sexes [@bolnick2003; @temeles2010]. Among penguins, bill morphology is a critical functional trait associated with foraging ecology, mate choice, and species recognition, making it a valuable feature for studying dimorphism [@gorman2014]. The palmerpenguins dataset [@horst2020] provides high-quality, standardized morphometric data for Adelie, Chinstrap, and Gentoo penguins from the Palmer Archipelago. These data allow for direct comparisons of bill length across sexes and species.

This study examines whether bill length consistently differs between males and females, and whether the magnitude of dimorphism varies among closely related species, potentially reflecting species-specific ecological pressures (e.g., diet, habitat). I hypothesized that males would exhibit longer bills than females within each species, with the degree of dimorphism differing by species. Analysis of the palmerpenguins dataset supports this hypothesis: males generally possess longer bills than females, with Chinstrap penguins showing the greatest mean bill length (\~51 mm), followed by Gentoo penguins (\~49 mm), while Adelie males have much shorter bills on average (\~40 mm).

# **Methods**

The palmerpenguins dataset was collected in the Palmer Archipelago (2007–2009) and released in a cleaned, teaching-friendly format (Horst et al., 2020; Gorman et al., 2014). A Two-way ANOVA with linear model was used to compared bill length differences between males and females across species. Data was analyzed in Rstudio using mainly the tidyverse package

# **Results**

```{r}
clean <- read.csv("../02_outdata/penguins_clean.csv")
means <- read.csv("../02_outdata/group_means_ci.csv")



means %>%
  arrange(species, sex) %>%
  knitr::kable(digits = 2,
               caption = "Mean bill length (±95% CI) by species and sex")
```


![Figure 1. Bill length by sex across Adelie, Chinstrap, and Gentoo penguins. Males generally exhibit longer bills than females, with the greatest differences